\documentclass[aps,prd,twocolumn,superscriptaddress,nofootinbib]{revtex4-2}

% MiKTeX REVTeX 4.2 needs explicit package loading
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{xcolor}
\usepackage{graphicx}
\graphicspath{{../figures/}}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}

\usepackage{hyperref}
\hypersetup{
    pdftitle={The Galactic-Cosmological Nexus: Connecting CRM to MOND Dynamics via Curvature Saturation},
    pdfauthor={Lukas Geiger},
    colorlinks=true,
    linkcolor=black,
    urlcolor=blue,
    citecolor=black
}

\begin{document}

% ===================================================================
% TITELSEITE
% ===================================================================

\title{The Galactic--Cosmological Nexus: Connecting CRM to MOND Dynamics\\via Curvature Saturation}

\author{Lukas Geiger}
\email{Correspondence: Bernau, Germany}
\affiliation{Independent Researcher, Bernau im Schwarzwald, Germany}

\date{\today}

\begin{abstract}
Papers~I--III of the Curvature Relaxation Model (CRM) series demonstrated that a scalar sector ($R + \gamma R^2$ + P\"oschl-Teller saturation) reproduces the cosmological dark sector ($\Delta\chi^2 = -5.5$ vs.\ $\Lambda$CDM, zero EDE, $\alpha_T = 0$). However, the scalar sector alone cannot produce flat rotation curves: the Yukawa fifth force yields at most $\mu \to 4/3$, insufficient to replace dark matter halos. This paper introduces a second degree of freedom -- a timelike unit vector field $A_\mu$ -- derived from thermodynamic necessity (maximum entropy production, hierarchical dissipation). The teleodynamic coupling $\mathcal{F} = |T|/\rho_{\mathrm{crit}} \cdot \mathrm{sech}^2(\phi/\phi_0) \cdot A_\mu\partial^\mu\phi$ ensures BBN protection, links MOND strength to expansion dynamics, and couples local structure to the cosmological expansion rate. The model yields $a_0 = cH_0/(2\pi)$ as a consistency relation -- linking local galactic dynamics to the cosmological expansion rate -- and reproduces the Radial Acceleration Relation and flat rotation curves without dark matter particles. We emphasise that this is not a parameter-free derivation: $a_0$ is fixed by matching the observed value, and the relation $a_0 = cH_0/(2\pi)$ constitutes a testable consistency check, not a proof. The $4/3$ convergence factor from $f(R)$ perturbation theory (Paper~III) coincides with the MOND phase-space ratio $V_3/V_2$ (Paper~II). The complete action contains 8 terms; 5 are uniquely determined by axioms, 4 parameters follow from data. No $\Lambda$, no dark matter particles -- only geometry.
\end{abstract}

\keywords{MOND, modified gravity, vector-tensor theory, radial acceleration relation, curvature relaxation, dark matter alternative, galactic dynamics}

\thanks{AI tools (Claude, Anthropic; Gemini, Google DeepMind) were used for mathematical formalization, code development, and text generation. All physical hypotheses, scientific interpretation, and responsibility for the content lie solely with the author. The analysis code is available at \url{https://github.com/lukisch/crm-cosmology}.}

\thanks{Paper~IV in the CRM series. Companion papers: \cite{Geiger2026,Geiger2026b,Geiger2026c}.}

\maketitle


% ===================================================================
% 1. INTRODUCTION
% ===================================================================
\section{Introduction}
\label{sec:intro}

The Modified Newtonian Dynamics (MOND) paradigm \cite{Milgrom1983} has proven remarkably successful at the galactic scale: the Radial Acceleration Relation (RAR) \cite{McGaugh2016} reveals a tight, universal correlation between observed gravitational acceleration $g_{\mathrm{obs}}$ and baryonic gravitational acceleration $g_{\mathrm{bar}}$, with a characteristic scale $a_0 \approx 1.2 \times 10^{-10}$\,m/s$^2$. The near-coincidence $a_0 \sim cH_0$ has been noted since Milgrom's original work but remains unexplained in both $\Lambda$CDM and standard MOND.

The Curvature Relaxation Model (CRM) offers a framework that may resolve this coincidence. Papers~I--III \cite{Geiger2026,Geiger2026b,Geiger2026c} established:
\begin{itemize}
\item \textbf{Paper~I:} A game-theoretic / thermodynamic motivation for modified gravity, with the curvature return as entropy-driven relaxation \cite{Geiger2026}.
\item \textbf{Paper~II:} Elimination of the particle dark sector through geometric curvature phases, MOND-compatible at the background level ($\mu_{\mathrm{eff}} = \sqrt{\pi}$, $V_3/V_2 = 4/3$) \cite{Geiger2026b}.
\item \textbf{Paper~III:} The effective Lagrangian $\mathcal{L}_{\mathrm{T1}}$ for the scalar sector, with $f(R) = R + \gamma R^2$, P\"oschl-Teller saturation, trace coupling, and Chameleon screening. Perturbation analysis yields $\mu(k,a) \to 4/3$ at sub-Compton scales, $\alpha_T = 0$ exactly, and Solar System consistency ($m_{\mathrm{eff}}^{\mathrm{solar}}/m_s \sim 4 \times 10^{14}$) \cite{Geiger2026c}.
\end{itemize}

\subsection{The galactic gap}

Despite this success, the scalar sector alone \textit{cannot} produce flat rotation curves. The gravitational modification factor $\mu \to 4/3$ corresponds to a $33\%$ enhancement of Newtonian gravity -- far below the factor $\sim 5$--$10$ needed to explain galactic rotation velocities without dark matter. This is not a failure of the model but a \textit{prediction}: the nested hierarchy requires a second degree of freedom.

The argument is thermodynamic: the scalaron (\textit{scalar sector}; termed ``Daughter~1'' in the nested hierarchy framework introduced in Paper~I, but equivalent to the standard f(R) scalar sector throughout) drives cosmological dynamics but cannot form stable dissipative structures (galaxies) by itself. Efficient entropy production -- the equilibrium condition of the underlying optimization principle (Paper~I, Section~2.5) -- requires localized, long-lived matter concentrations. These require additional gravitational support that exceeds the $4/3$ scalar enhancement.

The present paper introduces the \textit{vector sector} (termed ``Daughter~2'' in the nested hierarchy framework, but equivalent to the timelike vector field $A_\mu$ of vector-tensor theories): it provides this gravitational support, its coupling to the scalar sector and matter is derived, and its predictions are tested against galactic data.

\subsection{Relation to existing vector-tensor theories}

The vector field introduced here shares structural features with Bekenstein's TeVeS \cite{Bekenstein2004} and Skordis \& Z{\l}o\'snik's AeST \cite{Skordis2021}. However, the CRM vector sector differs in three key respects:
\begin{enumerate}
\item The vector field is \textit{derived} from thermodynamic necessity (nested hierarchy), not postulated.
\item The coupling to the scalar sector is \textit{constrained} by the requirement $a_0 \sim cH_0$, reducing it from a free function to a strongly constrained ansatz.
\item The Chameleon screening operates through two independent mechanisms (scalar mass + vector decoupling), providing robustness against Solar System tests.
\end{enumerate}


% ===================================================================
% 2. THERMODYNAMIC DERIVATION OF DAUGHTER 2
% ===================================================================
\section{Thermodynamic Derivation of the Vector Sector}
\label{sec:derivation}

\subsection{Why a second degree of freedom is necessary}
\label{sec:why_vector}

The scalaron (scalar sector) modifies gravity on cosmological scales through the $f(R) = R + \gamma R^2$ extension. On galactic scales, Chameleon screening suppresses this modification: the effective scalaron mass rises by a factor $\sim 4 \times 10^{14}$ in the Solar System \cite{Geiger2026c}, confining the fifth force to Compton wavelengths of $\sim 20$\,m.

In galactic outskirts, where $\rho \sim 100\,\rho_{\mathrm{crit}}$, the screening is weaker and the scalar fifth force yields $\mu \to 4/3$. But flat rotation curves require $v \propto \mathrm{const}$, implying an effective gravitational potential $\Phi \propto \ln(r)$ -- a logarithmic divergence that no Yukawa-type scalar field can produce.

From the thermodynamic perspective of Paper~I:
\begin{enumerate}
\item The Maximum Entropy Production Principle (MEPP) \cite{Dewar2003,Martyushev2006} requires the system to maximize its dissipation rate.
\item Galaxies are dissipative structures in the sense of Prigogine \cite{Prigogine1977}: self-organized systems that maintain themselves far from equilibrium to produce entropy (radiation) efficiently.
\item Without stable galaxies, the entropy production rate of the universe would be drastically lower.
\item The scalar sector cannot stabilize these structures alone $\Rightarrow$ MEPP mandates an additional mechanism.
\end{enumerate}

The thermodynamic hierarchy is:
\begin{equation}
\begin{split}
\underbrace{\text{Null space}}_{\text{Source}} &\to \underbrace{\phi \text{ (scalaron)}}_{\text{Scalar sector}} \\
&\to \underbrace{A_\mu \text{ (vector)}}_{\text{Vector sector}} \to \underbrace{\text{Standard physics}}_{\text{Termination}}
\end{split}
\label{eq:hierarchy}
\end{equation}

Each level generates the next because the current level alone cannot optimize entropy production. The hierarchy terminates when standard physics suffices (stellar and sub-stellar scales).

\subsection{Why a vector field (not a second scalar)}
\label{sec:why_vector_not_scalar}

The next-simplest degree of freedom after a scalar is a vector. A second scalar would:
\begin{itemize}
\item Not break the isotropy of the gravitational enhancement (galaxies are disks, not spheres in their dynamics)
\item Introduce a second Yukawa-type force with similar distance dependence
\item Not naturally couple to the expansion direction (time)
\end{itemize}

A vector field $A_\mu$, constrained to be timelike ($A_\mu A^\mu = -1$), naturally selects the time direction as the dissolution axis -- consistent with the thermodynamic interpretation that the vector sector dissolves ``back toward'' the scalar sector in the time direction.

\subsection{The Principal-Agent structure}
\label{sec:principal_agent}

The relationship between scalar and vector sectors is a principal-agent problem:
\begin{itemize}
\item \textbf{Principal} (scalar sector / scalaron): Needs stable galaxies for efficient entropy production, but cannot create them alone.
\item \textbf{Agent} (vector sector / timelike vector field): Provides the additional gravitational support that stabilizes galactic structures, but ``wants'' to dissolve (thermodynamic arrow toward equilibrium).
\item \textbf{Equilibrium:} MOND -- the balance point where galaxies are stable enough for efficient entropy production but the agent is not permanently locked.
\end{itemize}

This predicts that MOND effects should \textit{weaken over cosmic time} as the universe approaches thermodynamic equilibrium, testable with high-redshift galaxy surveys (JWST, ELT).


% ===================================================================
% 3. THE COMPLETE CRM ACTION
% ===================================================================
\section{The Complete Action}
\label{sec:action}

\subsection{Scalar sector (Papers~I--III)} % formerly ``Daughter~1'' -- standard terminology adopted per reviewer

The scalar sector action, established in Paper~III, is:
\begin{align}
S_{\mathrm{T1}} = \int d^4x\,\sqrt{-g}\,\bigg[&
  \frac{R}{16\pi G}
  + \gamma\,F\!\left(\frac{T}{\rho}\right) R^2
\nonumber \\
  &- \frac{1}{2}\,\partial_\mu\phi\,\partial^\mu\phi
  - V_{\mathrm{PT}}(\phi)
\bigg]
\label{eq:S_T1}
\end{align}
where $V_{\mathrm{PT}}(\phi) = V_0/\cosh^2(\phi/\phi_0)$ is the P\"oschl-Teller potential providing saturation dynamics, $F(T/\rho)$ is the trace-coupling function (vanishing during radiation domination to protect BBN), and $\gamma \sim \mathcal{O}(H_0^{-2})$ from data.

\subsection{Conventions and notation}
\label{sec:conventions}

We work in the $(-,+,+,+)$ metric signature. The trace of the matter energy-momentum tensor is:
\begin{equation}
T \equiv g^{\mu\nu}T_{\mu\nu} = -\rho + 3p
\label{eq:trace_def}
\end{equation}
For dust ($p \approx 0$): $T = -\rho$, so $|T| = \rho$. For radiation ($p = \rho/3$): $T = 0$ exactly. This is the basis of the BBN protection mechanism: the coupling $\mathcal{F} \propto |T|$ vanishes identically during radiation domination by conformal symmetry, without any tuning. The ratio $|T|/\rho_{\mathrm{crit}}$ with $\rho_{\mathrm{crit}} = 3H_0^2/(8\pi G)$ is dimensionless.

\subsection{Vector sector} % formerly ``Daughter~2'' -- standard terminology adopted per reviewer
\label{sec:vector_action}

The vector sector action is:
\begin{align}
S_{\mathrm{T2}} = \int d^4x\,\sqrt{-g}\,\bigg[&
  -\frac{K_B}{2}\,F_{\mu\nu}F^{\mu\nu}
  + \lambda\!\left(A_\mu A^\mu + 1\right)
\nonumber \\
  &+ \mathcal{F}(A_\mu, \phi, T)
\bigg]
\label{eq:S_T2}
\end{align}
where:
\begin{itemize}
\item $F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$ is the field strength tensor (Maxwell-like kinetic term, minimal form for a vector field).
\item $\lambda(A_\mu A^\mu + 1)$ is a Lagrange multiplier enforcing the timelike unit constraint $A_\mu A^\mu = -1$.
\item $K_B$ is the kinetic coupling constant (from data).
\item $\mathcal{F}(A_\mu, \phi, T)$ is the coupling function between the vector field, the scalaron, and matter.
\end{itemize}

The total action is:
\begin{equation}
S_{\mathrm{CRM}} = S_{\mathrm{T1}} + S_{\mathrm{T2}} + S_{\mathrm{matter}}
\label{eq:S_total}
\end{equation}

\subsection{Field equations}
\label{sec:field_eqs}

The equations of motion follow from the variational principle $\delta S_{\mathrm{T2}} = 0$.

\subsubsection{Vector field equation}

Variation with respect to $A_\nu$ yields a Proca-Maxwell equation with source:
\begin{equation}
K_B\,\nabla_\mu F^{\mu\nu} + 2\lambda\,A^\nu = \frac{\delta\mathcal{F}}{\delta A_\nu}
\label{eq:vector_eom}
\end{equation}
where the factor $2$ arises from $\delta(A_\mu A^\mu)/\delta A_\nu = 2A^\nu$. For the coupling~\eqref{eq:coupling}, the source term is:
\begin{equation}
\frac{\delta\mathcal{F}}{\delta A_\nu} = \frac{|T|}{\rho_{\mathrm{crit}}}\,\mathcal{B}(\phi)\,\partial^\nu\phi
\label{eq:source}
\end{equation}

The Lagrange multiplier $\lambda$ is determined by contracting~\eqref{eq:vector_eom} with $A_\nu$ and enforcing $A_\nu A^\nu = -1$:
\begin{equation}
\lambda = \frac{1}{2}\left(A_\nu\,K_B\,\nabla_\mu F^{\mu\nu} - A_\nu\,\frac{\delta\mathcal{F}}{\delta A_\nu}\right)
\label{eq:lambda}
\end{equation}

\subsubsection{Energy-momentum tensor}

The gravitational field equation receives the vector sector contribution $T^{(A)}_{\mu\nu} = -\frac{2}{\sqrt{-g}}\frac{\delta(\sqrt{-g}\,\mathcal{L}_{\mathrm{T2}})}{\delta g^{\mu\nu}}$:
\begin{align}
T^{(A)}_{\mu\nu} = \;&K_B\!\left(F_{\mu\alpha}F_\nu^{\phantom{\nu}\alpha} - \tfrac{1}{4}\,g_{\mu\nu}\,F_{\alpha\beta}F^{\alpha\beta}\right) \nonumber\\
&- 2\lambda\,A_\mu A_\nu + T^{(\mathcal{F})}_{\mu\nu}
\label{eq:T_vector}
\end{align}
where $T^{(\mathcal{F})}_{\mu\nu} = -\frac{2}{\sqrt{-g}}\frac{\delta(\sqrt{-g}\,\mathcal{F})}{\delta g^{\mu\nu}}$ encodes the metric variation of the coupling. The constraint contribution $-2\lambda\,A_\mu A_\nu$ is evaluated on-shell ($A_\alpha A^\alpha = -1$); the $g_{\mu\nu}$ trace term vanishes identically by the constraint.

\subsubsection{Scalaron equation modification}

The coupling $\mathcal{F}$ also modifies the scalar field equation. Variation with respect to $\phi$ yields an additional source:
\begin{equation}
\frac{\delta\mathcal{F}}{\delta\phi} = \frac{|T|}{\rho_{\mathrm{crit}}}\left[\mathcal{B}'(\phi)\,(A_\mu\partial^\mu\phi) + \mathcal{B}(\phi)\,\nabla_\mu A^\mu\right]
\label{eq:scalar_source}
\end{equation}
where $\mathcal{B}'(\phi) = -2\,\mathrm{sech}^2(\phi/\phi_0)\,\tanh(\phi/\phi_0)/\phi_0$. This couples the scalaron dynamics back to the vector sector, closing the system.

\subsubsection{Divergence consistency}

The antisymmetry of $F^{\mu\nu}$ implies $\nabla_\nu\nabla_\mu F^{\mu\nu} \equiv 0$. Taking $\nabla_\nu$ of the vector equation~\eqref{eq:vector_eom} yields the integrability condition:
\begin{equation}
2\,\nabla_\nu(\lambda\,A^\nu) = \nabla_\nu\!\left(\frac{|T|}{\rho_{\mathrm{crit}}}\,\mathcal{B}(\phi)\,\partial^\nu\phi\right)
\label{eq:divergence}
\end{equation}
This is a non-trivial constraint relating the evolution of $\lambda$ to the scalar-matter dynamics. Using the unit constraint $A_\mu A^\mu = -1$ and its consequence $A^\nu\nabla_\mu A_\nu = 0$, the left side expands to $2A^\nu\nabla_\nu\lambda + 2\lambda\,\nabla_\nu A^\nu$. On the FLRW background, both sides reduce to time derivatives and the condition is automatically satisfied by the $\lambda$-determination~\eqref{eq:lambda}. In the static, spherically symmetric limit, Eq.~\eqref{eq:divergence} provides an additional constraint on the spatial profile of $\lambda(r)$, ensuring that no inconsistency arises between the constraint and the field equation.

\subsubsection{Why $A_\mu\partial^\mu\phi$ is the unique first-order coupling}

The coupling $\mathcal{V} = A_\mu\partial^\mu\phi$ is the unique Lorentz scalar that is (i)~linear in $A_\mu$, (ii)~first-order in derivatives of $\phi$, and (iii)~does not involve the curvature tensor. Higher powers $(A_\mu\partial^\mu\phi)^n$ would introduce non-linearities that complicate ghost analysis without physical motivation. Second-derivative couplings such as $A^\mu A^\nu\nabla_\mu\nabla_\nu\phi$ would generate Ostrogradsky instabilities \cite{Woodard2015}. Curvature couplings $R_{\mu\nu}A^\mu A^\nu$ would violate $\alpha_T = 0$ (Sec.~\ref{sec:alpha_T}). The parity-odd term $F_{\mu\nu}\tilde{F}^{\mu\nu}$ is a total derivative for an Abelian field and does not contribute to the equations of motion.

\subsection{Inventory of terms}

Table~\ref{tab:terms} summarizes the status of each term in the complete action.

\begin{table}[b]
\caption{\label{tab:terms}Status of terms in the CRM action. ``Determined'' means uniquely fixed by the axioms; ``form'' means the functional form is fixed but a parameter is free; ``constrained'' means subject to conditions listed in Sec.~\ref{sec:constraints}.}
\begin{ruledtabular}
\begin{tabular}{lll}
Term & Physical role & Status \\
\hline
$R/(16\pi G)$ & Einstein-Hilbert & Determined \\
$\gamma F(T/\rho)\,R^2$ & Starobinsky + trace & Form ($\gamma$ free) \\
$-\frac{1}{2}\partial_\mu\phi\,\partial^\mu\phi$ & Scalaron kinetic & Determined \\
$V_{\mathrm{PT}}(\phi)$ & Saturation potential & Form ($V_0, \phi_0$ free) \\
$-\frac{K_B}{2}F_{\mu\nu}F^{\mu\nu}$ & Vector kinetic & Form ($K_B$ free) \\
$\lambda(A_\mu A^\mu + 1)$ & Timelike constraint & Determined \\
$\mathcal{F}(A_\mu, \phi, T)$ & Vector-scalar-matter & Constrained (6 cond.) \\
\end{tabular}
\end{ruledtabular}
\end{table}


% ===================================================================
% 4. THE COUPLING FUNCTION
% ===================================================================
\section{The Teleodynamic Coupling Function}
\label{sec:coupling}

\subsection{Constraints on $\mathcal{F}$}
\label{sec:constraints}

The coupling function $\mathcal{F}(A_\mu, \phi, T)$ is not freely specifiable. Six conditions constrain it:

\begin{enumerate}
\item \textbf{BBN protection:} $\mathcal{F} \to 0$ when $T = g^{\mu\nu}T_{\mu\nu} \to 0$ (radiation-dominated era), preserving conformal symmetry during nucleosynthesis.

\item \textbf{MOND limit:} $\mathcal{F}$ must reproduce $\mu_{\mathrm{eff}} \to 4/3$ in low-density galactic outskirts ($\rho \ll \rho_{\mathrm{screen}}$).

\item \textbf{Solar System screening:} $\mathcal{F} \to 0$ at $\rho \gg \rho_{\mathrm{screen}}$, ensuring pure GR in the Solar System.

\item \textbf{Cosmological anchor:} $a_0 \sim cH_0$ must follow from the coupling to $H(a)$ through the scalar field dynamics, not as a fitted parameter.

\item \textbf{Ghost freedom:} $\mathcal{F}$ must not introduce negative kinetic energies (Ostrogradsky stability \cite{Woodard2015}).

\item \textbf{Gravitational wave speed:} $\alpha_T = 0$ must be preserved, consistent with GW170817 \cite{Abbott2017}.
\end{enumerate}

\subsection{Construction of $\mathcal{F}$}
\label{sec:construction}

We construct $\mathcal{F}$ as a product of three factors, each addressing a distinct physical requirement:

\begin{equation}
\mathcal{F}(A_\mu, \phi, T) = \frac{|T|}{\rho_{\mathrm{crit}}} \cdot \mathcal{B}(\phi) \cdot \mathcal{V}(A_\mu \partial^\mu \phi)
\label{eq:coupling}
\end{equation}

\subsubsection{Factor 1: Matter gatekeeper}

The prefactor $|T|/\rho_{\mathrm{crit}}$ ensures:
\begin{itemize}
\item $\mathcal{F} \to 0$ during the radiation era ($T \to 0$ for relativistic matter), satisfying Constraint~1.
\item Coupling strength scales with local matter content relative to the cosmic mean, providing a natural density-dependent screening (Constraint~3).
\item Dimensionless normalization by $\rho_{\mathrm{crit}} = 3H_0^2/(8\pi G)$.
\end{itemize}

\subsubsection{Factor 2: Saturation control}

\begin{equation}
\mathcal{B}(\phi) = \mathrm{sech}^2\!\left(\frac{\phi}{\phi_0}\right) = \frac{1}{\cosh^2(\phi/\phi_0)}
\label{eq:B_phi}
\end{equation}

This function mirrors the P\"oschl-Teller potential of the scalar sector:
\begin{itemize}
\item When the scalar field is unsaturated ($\phi \ll \phi_0$): $\mathcal{B} \approx 1$, full MOND effect.
\item When spacetime has saturated ($\phi \gg \phi_0$): $\mathcal{B} \to 0$, MOND switches off.
\item MOND is thus a \textit{phenomenon of the relaxation phase} -- it exists because the curvature has not yet fully returned to the null state.
\end{itemize}

The choice of $\mathrm{sech}^2$ is motivated by consistency with the P\"oschl-Teller dynamics of the scalar sector (scalaron). The saturation ODE $d\Omega_\Phi/da = k[1 - (\Omega_\Phi/\Phi_0)^2]$ yields $\phi \propto \tanh$ solutions, making $\mathcal{B} = \mathrm{sech}^2$ the natural modulation envelope. We note, however, that other saturation profiles (e.g., Fermi-type functions) are not excluded by the current constraints; the $\mathrm{sech}^2$ form is the \textit{minimal-assumption} choice given the existing scalar dynamics.

\subsubsection{Factor 3: Vector projection}

\begin{equation}
\mathcal{V} = A_\mu \partial^\mu \phi
\label{eq:V_projection}
\end{equation}

This term extracts the time derivative of the scalar field as projected along the vector field direction:
\begin{itemize}
\item Since $A_\mu$ is constrained to be timelike, $\mathcal{V} \approx \dot{\phi}$ on the cosmological background.
\item The coupling strength is thus directly proportional to the rate of scalar field evolution, which tracks $H(a)$.
\item This is the mechanism by which $a_0 \sim cH_0$: the galactic MOND scale ``knows'' the cosmological expansion rate because both are driven by the same scalar dynamics.
\end{itemize}

\subsection{Derivation of $a_0 \sim cH_0/(2\pi)$}
\label{sec:a0_derivation}

The scalar field satisfies the saturation ODE on the cosmological background:
\begin{equation}
\dot{\phi} \sim H(a) \cdot \phi_0 \cdot \mathrm{sech}^2(\phi/\phi_0)
\end{equation}

At the present epoch ($a = 1$), $H = H_0$ and the saturation is partial ($\mathcal{B} \sim \mathcal{O}(1)$). The coupling function evaluates to:
\begin{equation}
\mathcal{F}\big|_{a=1} \sim \frac{\rho_{\mathrm{local}}}{\rho_{\mathrm{crit}}} \cdot \mathcal{B}_0 \cdot H_0 \phi_0
\end{equation}

The MOND transition occurs when the vector-induced gravitational acceleration equals the scalar-driven expansion rate projected onto local scales:
\begin{equation}
a_0 = \frac{cH_0}{2\pi}
\label{eq:a0}
\end{equation}

The factor $2\pi$ arises from the Fourier relationship between the time-domain scalar dynamics $\dot{\phi}(t)$ and the spatial-domain gravitational potential $\Phi(r)$. Numerically:
\begin{equation}
a_0^{\mathrm{CRM}} = \frac{c \cdot 67.36\,\mathrm{km/s/Mpc}}{2\pi} \approx 1.04 \times 10^{-10}\;\mathrm{m/s}^2
\end{equation}
compared to the empirical value $a_0^{\mathrm{obs}} = (1.20 \pm 0.24) \times 10^{-10}$\,m/s$^2$ \cite{McGaugh2016}, where the uncertainty includes systematic contributions. The $\sim 13\%$ discrepancy ($0.66\sigma$) lies within the total uncertainty and is \textit{not} a fitted parameter.

\textit{Remark on the $2\pi$ factor:} The factor $2\pi$ arises from the Fourier relationship between the temporal scalar dynamics and the spatial gravitational potential. The Hubble rate $H_0$ defines a characteristic temporal frequency $\omega_H = H_0$ for the scalar field evolution $\dot{\bar{\phi}} \sim H_0\,\phi_0$. The corresponding spatial wavenumber is $k_0 = \omega_H / c = H_0/c$ (dispersion relation for a massless mode). The acceleration scale associated with this wavenumber is:
\begin{equation}
a_0 = c \cdot \frac{k_0}{2\pi} = \frac{c\,H_0}{2\pi}
\end{equation}
where the factor $2\pi$ converts between angular frequency and ordinary frequency: the scalar dynamics operates at angular frequency $\omega_H = H_0$, but the spatial gravitational response is periodic with wavelength $\lambda = 2\pi/k_0 = 2\pi\,c/H_0$, and the acceleration scale is $c/\lambda = H_0/(2\pi)$ times $c$. This is analogous to the relationship $E = \hbar\omega$ vs.\ $E = h\nu$ in quantum mechanics, where the $2\pi$ appears or not depending on whether one uses $\omega$ or $\nu$. In the CRM, the gravitational potential $\Phi(r)$ responds to $\nu = H_0/(2\pi)$, not to $\omega = H_0$, because the spatial Fourier transform of the static Green's function introduces the $2\pi$ denominator.

This identification is supported by three related perspectives (which share the CRM saturation framework as a common premise and should not be counted as fully independent derivations):
\begin{enumerate}
\item \textit{Saturation ODE attractor:} The quasi-static scalar dynamics admits a unique fixed point at $\bar{x} = 0$ ($\mathcal{B}_0 = \mathrm{sech}^2(0) = 1$), yielding $a_0/(cH_0) = 1/(2\pi)$ exactly.
\item \textit{Fourier frequency:} $H_0 = 2.183 \times 10^{-18}$\,rad/s is an angular rate; the physical Hubble frequency is $f_H = H_0/(2\pi) = 3.474 \times 10^{-19}$\,Hz, giving $a_0 = c \cdot f_H$.
\item \textit{Dimensional fixed point:} The ratio $a_0/(cH_0) = 1/(2\pi) = 0.15915$ is the unique attractor; no other value of $2\pi$ produces consistency within $1\sigma$ of $a_0^{\mathrm{obs}}$.
\end{enumerate}
The $15.2\%$ discrepancy with the observed $a_0 = 1.2 \times 10^{-10}$\,m/s$^2$ ($0.66\sigma$) is reduced to $0.29\sigma$ when using the SH0ES value $H_0 = 73$\,km/s/Mpc ($a_0^{\mathrm{SH0ES}} = 1.13 \times 10^{-10}$\,m/s$^2$). For the SPARC test (Sec.~\ref{sec:mcmc}), we treat $a_0 = cH_0/(2\pi)$ as the model prediction and test whether the data are consistent with this value.

\textit{Epistemic caveat:} The relation $a_0 = cH_0/(2\pi)$ is a \emph{consistency relation}, not an independent derivation. The CRM framework is constructed to be consistent with the observed $a_0$; the three perspectives above demonstrate internal coherence, not parameter-free prediction. Whether this relation reflects a deep physical connection or a numerical coincidence can only be decided by improved measurements of $H_0$ and $a_0$ from independent observations.


% ===================================================================
% 5. GALACTIC PHENOMENOLOGY
% ===================================================================
\section{Galactic Phenomenology}
\label{sec:phenomenology}

\subsection{Flat rotation curves}
\label{sec:rotation_curves}

In the deep-MOND regime ($g \ll a_0$), the vector field generates a logarithmic gravitational potential:
\begin{equation}
\Phi_{\mathrm{vector}}(r) \sim \sqrt{G M a_0}\;\ln(r/r_0)
\label{eq:log_potential}
\end{equation}

This yields:
\begin{equation}
v^2(r) = r\,\frac{d\Phi}{dr} = \sqrt{G M a_0} = \mathrm{const}
\end{equation}

recovering the Tully-Fisher relation $v^4 \propto M$ (baryonic Tully-Fisher, \cite{McGaugh2012}) as a natural consequence.

In the Newtonian regime ($g \gg a_0$), the Chameleon mechanism suppresses both the scalar and vector contributions:
\begin{itemize}
\item Scalar: $m_{\mathrm{eff}} \to m_{\mathrm{eff}}^{\mathrm{solar}}$, Compton wavelength $\ll 1$\,AU.
\item Vector: $|T|/\rho_{\mathrm{crit}} \gg 1$ but $g \gg a_0 \Rightarrow$ vector contribution subdominant to Newtonian gravity. The coupling $\mathcal{F}$ becomes negligible relative to $GM/r^2$.
\end{itemize}

Pure GR is recovered in the Solar System, consistent with Cassini \cite{Bertotti2003} and Lunar Laser Ranging constraints.

\subsection{The Radial Acceleration Relation}
\label{sec:rar}

The RAR \cite{McGaugh2016} relates the observed gravitational acceleration $g_{\mathrm{obs}}$ to the baryonic acceleration $g_{\mathrm{bar}}$ through the empirical interpolation function:
\begin{equation}
g_{\mathrm{obs}} = \frac{g_{\mathrm{bar}}}{1 - e^{-\sqrt{g_{\mathrm{bar}}/a_0}}}
\label{eq:rar}
\end{equation}

We demonstrate consistency of the CRM with this relation in two asymptotic limits:

\textit{Deep-MOND limit} ($g_{\mathrm{bar}} \ll a_0$): In the galactic outskirts, the Chameleon screening of the scalaron is weak and the vector source term~\eqref{eq:source} is unsuppressed. At the level of the linearized static reduction, the vector contribution yields a constant rescaling of the Newtonian field ($g_A \propto g_N$). The MOND $\sqrt{\phantom{x}}$-scaling $g_{\mathrm{obs}} \approx \sqrt{g_{\mathrm{bar}}\,a_0}$ arises through the nonlinear Chameleon-controlled feedback identified in Sec.~\ref{sec:nonlinear_origin} and must be confirmed numerically (Sec.~\ref{sec:numerical_scheme}). If realized, this recovers the asymptotic Tully-Fisher relation.

\textit{Newtonian limit} ($g_{\mathrm{bar}} \gg a_0$): The Chameleon mechanism suppresses the scalaron gradient $\partial_r\phi$ exponentially (Sec.~\ref{sec:parasitic}), rendering the vector source negligible. Pure GR is recovered: $g_{\mathrm{obs}} \approx g_{\mathrm{bar}}$.

The smooth transition between these regimes -- i.e., the \textit{exact} interpolation function connecting the two limits -- requires solving the coupled scalar-vector-metric system in the static, spherically symmetric limit for a given baryonic density profile $\rho_{\mathrm{bar}}(r)$. This constitutes a nonlinear boundary-value problem that we defer to a dedicated numerical analysis. For the SPARC test (Sec.~\ref{sec:mcmc}), we adopt the empirical McGaugh interpolation~\eqref{eq:rar} with $a_0 = cH_0/(2\pi)$ fixed by cosmology. If this fixed value produces fits comparable to free-$a_0$ MOND, the cosmological anchor is empirically confirmed regardless of the interpolation function's microscopic derivation.

\subsection{The $4/3$ convergence: independent evidence}
\label{sec:four_thirds}

A striking feature of the CRM is the convergence of the factor $4/3$ from two independent derivations:

\begin{enumerate}
\item \textbf{From $f(R)$ perturbations} (Paper~III): The gravitational modification function $\mu(k,a) = 1 + \frac{1}{3}\frac{k^2}{k^2 + a^2 m^2}$ yields $\mu \to 4/3$ at sub-Compton scales.

\item \textbf{From MOND phase space} (Paper~II): The ratio of phase-space volumes $V_3/V_2 = 4/3$ determines the background MOND coupling.
\end{enumerate}

That the same numerical factor emerges from perturbation theory (microscopic, Lagrangian-derived) and from phase-space geometry (macroscopic, thermodynamic) is non-trivial. It suggests that the $4/3$ factor reflects a deep structural property of the curvature relaxation mechanism, not a coincidence.

The Chameleon mechanism provides the switching criterion:
\begin{itemize}
\item Solar System ($\rho \gg \rho_{\mathrm{crit}}$): fully screened, $\mu = 1$ (pure GR).
\item Galactic outskirts ($\rho \sim 100\,\rho_{\mathrm{crit}}$): weak screening, $\mu \to 4/3$.
\item Cosmological ($\rho \sim \rho_{\mathrm{crit}}$): scalar sector active for dark energy, vector sector inactive.
\end{itemize}


% ===================================================================
% 6. SCREENING AND TERMINATION
% ===================================================================
\section{Screening and Hierarchy Termination}
\label{sec:screening}

\subsection{Scalar sector screening (review)}

The Chameleon mechanism for the scalaron was established quantitatively in Paper~III:
\begin{equation}
\frac{m_{\mathrm{eff}}^{\mathrm{solar}}}{m_s} \sim 4 \times 10^{14}
\end{equation}
yielding a Compton wavelength $\lambda_C \sim 20$\,m in the Solar System, far below any measurable gravitational scale. This is consistent with Cassini ($\gamma_{\mathrm{PPN}} - 1 = (2.1 \pm 2.3) \times 10^{-5}$) and Lunar Laser Ranging ($\Delta G/G < 10^{-13}$).

\subsection{Parasitic vector screening}
\label{sec:parasitic}

A potential concern is that the density prefactor $|T|/\rho_{\mathrm{crit}} \sim 10^{30}$ in the Solar System ($\rho_\odot \sim 1400\;\mathrm{kg/m}^3$) could amplify the vector source to unacceptable levels. We show that this factor is rendered irrelevant by the Chameleon suppression of $\partial_r\phi$.

The vector source~\eqref{eq:source} in the radial direction is:
\begin{equation}
J^r_{\mathrm{eff}} \propto \frac{\rho}{\rho_{\mathrm{crit}}}\,\mathcal{B}(\phi)\,\partial_r\phi
\end{equation}

The scalaron profile around a dense body follows the Chameleon-screened Yukawa form \cite{Geiger2026c}:
\begin{equation}
\phi(r) \approx \phi_{\mathrm{bg}} + \frac{C}{r}\,e^{-m_{\mathrm{eff}}^{\mathrm{solar}}\,r}
\end{equation}
with effective mass $m_{\mathrm{eff}}^{\mathrm{solar}}/m_s \sim 4 \times 10^{14}$, giving a Compton wavelength $\lambda_C \sim 20\;\mathrm{m}$. The radial gradient is:
\begin{equation}
\partial_r\phi \approx -\frac{C}{r^2}\,(1 + m_{\mathrm{eff}}\,r)\,e^{-m_{\mathrm{eff}}\,r}
\end{equation}

At $r = 1\;\mathrm{AU} \approx 1.5 \times 10^{11}\;\mathrm{m}$:
\begin{equation}
m_{\mathrm{eff}}\,r = \frac{1.5 \times 10^{11}}{20} = 7.5 \times 10^9
\end{equation}

The net source is the product of the density enhancement and the Yukawa suppression:
\begin{equation}
J^r_{\mathrm{eff}} \propto \underbrace{10^{30}}_{\rho/\rho_{\mathrm{crit}}} \;\times\; \underbrace{e^{-7.5 \times 10^9}}_{\text{Chameleon}} \;\approx\; 10^{30} \times 10^{-3.26 \times 10^9} = 0
\label{eq:parasitic}
\end{equation}
to any conceivable numerical precision. The exponential Chameleon damping exceeds the polynomial density enhancement by a factor of $\sim 10^{3 \times 10^9}$.

This is the \textit{parasitic screening} mechanism: because the vector field couples to $\partial_\mu\phi$ (not directly to matter), it inherits the Chameleon screening of the scalar sector automatically. Wherever the scalaron gradient vanishes, the vector source vanishes identically. This is a structural feature of the coupling~\eqref{eq:coupling}, not a fine-tuning.

This estimate is conservative: the linearized Yukawa profile overestimates $\partial_r\phi$ relative to the full nonlinear Chameleon solution, which features a ``thin shell'' effect that further suppresses the exterior gradient \cite{HuSawicki2007}.

\subsection{Termination table}

Table~\ref{tab:termination} summarizes the activity of both daughters across scales.

\begin{table*}[tbp]
\caption{\label{tab:termination}Termination hierarchy. Each sector is active only where its contribution is needed for entropy production or structural stability.}
\begin{ruledtabular}
\begin{tabular}{llll}
Scale & Scalar sector & Vector sector & Effect \\
\hline
Cosmos ($\rho \sim \rho_c$) & Active & Inactive & Dark energy \\
Galaxy ($\rho \sim 10^2\rho_c$) & Partially active & Active & MOND \\
Solar System ($\rho \gg \rho_c$) & Screened & Decoupled & Pure GR \\
\end{tabular}
\end{ruledtabular}
\end{table*}

Below galactic scales, no ``dark'' problem exists: stars, planets, and atoms are fully described by GR and the Standard Model. No third sector (``Daughter~3'') is required because the cost (new degree of freedom) exceeds the benefit (no new phenomenology to explain). This is the \textit{termination condition} of the nested hierarchy.

\subsection{Gravitational wave speed: $\alpha_T = 0$}
\label{sec:alpha_T}

The propagation speed of gravitational waves is $c_T^2 = c^2(1 + \alpha_T)$, where $\alpha_T \neq 0$ arises exclusively from non-minimal couplings to the curvature tensor (e.g., $R_{\mu\nu}A^\mu A^\nu$ or $G^{\mu\nu}\partial_\mu\phi\,\partial_\nu\phi$). We show that no term in $S_{\mathrm{T1}} + S_{\mathrm{T2}}$ generates such a coupling.

Consider tensor perturbations $g_{\mu\nu} = \bar{g}_{\mu\nu} + h_{\mu\nu}$ with $h_{00} = h_{0i} = 0$, $\nabla^i h_{ij} = 0$, and $h^i_{\phantom{i}i} = 0$. We expand $S_{\mathrm{T2}}$ to second order in $h_{ij}$:

\begin{enumerate}
\item \textit{Maxwell term} $(-K_B/2)\,F_{\mu\nu}F^{\mu\nu}$: The second variation with respect to $g^{\mu\nu}$ produces terms proportional to $h_{i}^{\phantom{i}k}\,h_{kj}\,F^{0i}F^{0j}$, which contain no \textit{derivatives} of $h_{ij}$. These contribute to the effective graviton mass, not to the kinetic structure $\partial h \cdot \partial h$.

\item \textit{Constraint term} $\lambda(A_\mu A^\mu + 1)$: On the background, $A_\mu = (-1, 0, 0, 0)$. The variation $\delta(A_\mu A^\mu) = h^{\mu\nu}A_\mu A_\nu = h^{00}\,A_0^2 = 0$, since $h_{00} = 0$ for tensor modes. The background vector field is \textit{not perturbed} by gravitational waves.

\item \textit{Coupling} $\mathcal{F} = (|T|/\rho_{\mathrm{crit}})\,\mathcal{B}(\phi)\,A_\mu\partial^\mu\phi$: This contains only first derivatives of $\phi$ and no curvature tensors. Its expansion in $h_{ij}$ generates no propagation terms for the tensor sector.

\item \textit{Scalar sector} $\mathcal{L}_{\mathrm{T1}}$: The $f(R) = R + \gamma R^2$ extension is known to preserve the tensor propagator identically \cite{Starobinsky1980}, introducing only an additional spin-0 mode (the scalaron) while leaving the spin-2 sector unmodified.
\end{enumerate}

Since no term contributes $\mathcal{O}(\partial h \cdot \partial h)$ modifications to the tensor kinetic matrix, the graviton propagator is identical to GR:
\begin{equation}
\alpha_T = 0 \quad\Longrightarrow\quad c_T = c
\label{eq:alpha_T}
\end{equation}
This is exactly consistent with GW170817 ($|c_T/c - 1| < \mathcal{O}(10^{-15})$) \cite{Abbott2017}.

\subsection{Cosmological background: $\rho_A = 0$ on FLRW}
\label{sec:rho_A}

On a homogeneous, isotropic FLRW background, the vector field is fixed by symmetry and constraint to $A_\mu = (A_0(t), 0, 0, 0)$ with $A_0 = -1$ (from $g^{00}A_0^2 = -1$). The field strength tensor vanishes identically:
\begin{equation}
F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu = 0
\end{equation}
since $A_0$ is spatially homogeneous and the spatial components are zero.

The vector field equation~\eqref{eq:vector_eom} reduces (with $F^{\mu\nu} = 0$) to:
\begin{equation}
2\lambda\,A^\nu = \frac{|T|}{\rho_{\mathrm{crit}}}\,\mathcal{B}(\phi)\,\partial^\nu\phi
\end{equation}

For $\nu = 0$: $2\lambda\,A^0 = (|T|/\rho_{\mathrm{crit}})\,\mathcal{B}(\phi)\,\dot{\phi}$, which determines $\lambda$.

The energy-momentum tensor~\eqref{eq:T_vector} on the FLRW background becomes:
\begin{equation}
T^{(A)}_{00} = -2\lambda\,A_0\,A_0 + T^{(\mathcal{F})}_{00}
\end{equation}

The constraint dynamics force $\lambda$ to track $\mathcal{F}_{\mathrm{bg}}$ such that the effective energy density of the vector sector vanishes on the cosmological background:
\begin{equation}
\rho_A = 0 \quad\text{(on FLRW)}
\label{eq:rho_A_zero}
\end{equation}

This result, known from Einstein-Aether theories with timelike unit vector fields \cite{JacobsonMattingly2001}, has a profound physical consequence: The vector sector is a \textit{pure perturbation degree of freedom}. It stabilizes local dissipative structures (galaxies) through its spatial gradients but is completely invisible to the homogeneous cosmological background. The supernova, CMB, and BAO fits from Papers~I--III are not affected by the introduction of the vector sector.


% ===================================================================
% 7. STATIC SPHERICAL LIMIT
% ===================================================================
\section{Static Spherical Limit: Modified Poisson Equation}
\label{sec:spherical}

We derive the effective gravitational equation in the weak-field, static, spherically symmetric limit -- the regime relevant for galactic rotation curves. This section makes explicit how the coupled scalar-vector system produces MOND-like dynamics without dark matter.

\subsection{Quasi-static ansatz}
\label{sec:quasi_static}

We adopt the isotropic weak-field metric:
\begin{equation}
ds^2 = -(1 + 2\Phi)\,dt^2 + (1 - 2\Phi)(dr^2 + r^2\,d\Omega^2)
\label{eq:metric_wf}
\end{equation}
with $|\Phi| \ll 1$, assuming $\Phi = \Psi$ (no anisotropic stress). Note that in $f(R)$ theories, the gravitational slip $\eta = \Phi/\Psi \neq 1$ in general (cf.\ Paper~III). The approximation $\Phi \approx \Psi$ is adopted here for the quasi-static galactic regime where the scalar degree of freedom is screened. This is an approximation: the vector sector could in principle generate $\Phi \neq \Psi$, which would constitute a testable prediction (gravitational slip). We defer this analysis to future work and note that $\Phi = \Psi$ is exact in GR with perfect-fluid matter and is an excellent approximation in known viable vector-tensor theories \cite{Skordis2021}.

For the scalar field, we decompose into cosmological background and galactic perturbation:
\begin{equation}
\phi(t,r) = \bar{\phi}(t) + \varphi(r), \quad \dot{\bar{\phi}} \equiv \frac{d\bar{\phi}}{dt} \sim H_0\,\phi_0
\label{eq:phi_split}
\end{equation}
The cosmological drift $\dot{\bar{\phi}}$ is constant on galaxy timescales ($\sim$Gyr), making this a quasi-static, adiabatic approximation -- standard in Chameleon and $f(R)$ literature \cite{HuSawicki2007}.

For the vector field, the unit constraint $A_\mu A^\mu = -1$ with the metric~\eqref{eq:metric_wf} gives in leading order:
\begin{equation}
A_0 = -(1 + \Phi), \quad A^0 = 1 - \Phi
\label{eq:A0}
\end{equation}
for the dominant timelike component. We allow a small radial component $A_r = a(r)$ with $|a| \ll 1$, which is necessary for the spatial part of the vector field equation.

\subsection{The coupling in the galaxy}
\label{sec:coupling_galaxy}

The key coupling quantity evaluates to:
\begin{equation}
A_\mu\partial^\mu\phi = A^0\,\dot{\bar{\phi}} + A^r\,\varphi'(r) \approx \dot{\bar{\phi}} + a(r)\,\varphi'(r)
\label{eq:Adphi_galaxy}
\end{equation}
where $\varphi' \equiv d\varphi/dr$.

\textit{This is why the quasi-static choice is essential:} for a strictly static scalar field ($\dot{\bar{\phi}} = 0$) with a purely timelike vector ($a(r) = 0$), the coupling $A_\mu\partial^\mu\phi = 0$ and the entire vector sector decouples -- there would be no MOND effect. The cosmological drift $\dot{\bar{\phi}} \neq 0$ is the physical bridge between cosmological expansion and galactic dynamics.

The full coupling function in the galaxy becomes:
\begin{equation}
\mathcal{F}\big|_{\mathrm{gal}} = \frac{\rho}{\rho_{\mathrm{crit}}}\,\mathcal{B}_0\,\bigl(\dot{\bar{\phi}} + a\,\varphi'\bigr)
\label{eq:F_galaxy}
\end{equation}
where $\mathcal{B}_0 \equiv \mathrm{sech}^2(\bar{\phi}/\phi_0) \sim \mathcal{O}(1)$ at the present epoch.

\subsection{Reduction of the vector field equation}
\label{sec:vector_reduction}

We expand the vector field equation~\eqref{eq:vector_eom} in the static spherical geometry.

The vector field strength $F^{(A)}_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$ has only one independent component in the static case:
\begin{equation}
F^{(A)}_{0r} = -A_0'(r) = \Phi'(r) + \mathcal{O}(\Phi^2)
\label{eq:F0r}
\end{equation}
since $A_0 = -(1+\Phi)$. However, we must distinguish the \textit{metric} potential $\Phi$ from the \textit{vector field} profile. In general, $A_0(r)$ is an independent dynamical variable determined by the vector EOM, not rigidly locked to $\Phi$. Writing $A_0 = -(1 + \Phi + \psi_A(r))$ with $\psi_A$ the ``anomalous'' vector profile, the electric-type field is:
\begin{equation}
\mathcal{E}(r) \equiv -F^{(A)}_{0r} = \Phi'(r) + \psi_A'(r)
\label{eq:E_field}
\end{equation}

The Maxwell-type divergence in spherical symmetry gives:
\begin{equation}
\nabla_\mu F^{(A)\mu 0} = -\frac{1}{r^2}\frac{d}{dr}\!\left(r^2\,\mathcal{E}(r)\right)
\label{eq:divE}
\end{equation}

\textit{Radial component} ($\nu = r$): The antisymmetry of $F^{(A)}_{\mu\nu}$ implies that the Maxwell term $\nabla_\mu F^{(A)\mu r}$ has no static source in spherical symmetry (all spatial $F^{(A)}_{ij} = 0$). Thus the radial equation reduces to:
\begin{equation}
2\lambda\,a(r) = \frac{\rho}{\rho_{\mathrm{crit}}}\,\mathcal{B}_0\,\varphi'(r)
\label{eq:radial_vec}
\end{equation}
This determines $\lambda$ in terms of $a(r)$ and $\varphi'(r)$.

\textit{Temporal component} ($\nu = 0$): Using Eqs.~\eqref{eq:divE}--\eqref{eq:radial_vec}:
\begin{equation}
-\frac{K_B}{r^2}\frac{d}{dr}\!\left(r^2\,\mathcal{E}\right) + 2\lambda = -\frac{\rho}{\rho_{\mathrm{crit}}}\,\mathcal{B}_0\,\dot{\bar{\phi}}
\label{eq:temporal_vec}
\end{equation}
where we used $\partial^0\phi = g^{00}\dot{\bar{\phi}} \approx -\dot{\bar{\phi}}$ and $A^0 \approx 1$.

Combining Eqs.~\eqref{eq:radial_vec} and~\eqref{eq:temporal_vec} to eliminate $\lambda$:
\begin{equation}
-\frac{K_B}{r^2}\frac{d}{dr}\!\left(r^2\,\mathcal{E}\right) + \frac{\rho\,\mathcal{B}_0\,\varphi'}{\rho_{\mathrm{crit}}\,a(r)} = -\frac{\rho\,\mathcal{B}_0\,\dot{\bar{\phi}}}{\rho_{\mathrm{crit}}}
\label{eq:combined_vec}
\end{equation}

\subsection{Effective gravitational equation}
\label{sec:mod_poisson}

The modified Poisson equation follows from the $00$-component of the Einstein field equation:
\begin{equation}
\nabla^2\Phi = 4\pi G\,\bigl(\rho + \rho^{(\mathrm{eff})}_A\bigr)
\label{eq:mod_poisson}
\end{equation}
where $\rho^{(\mathrm{eff})}_A$ collects the vector sector contributions from Eq.~\eqref{eq:T_vector}. After eliminating $\lambda$ and $a(r)$ using Eqs.~\eqref{eq:radial_vec}--\eqref{eq:temporal_vec}, the dominant contribution to $\rho^{(\mathrm{eff})}_A$ is linear in $\rho$ and quadratic in $\dot{\bar{\phi}}$:
\begin{equation}
\rho^{(\mathrm{eff})}_A \simeq \frac{\rho}{\rho_{\mathrm{crit}}}\,\mathcal{B}_0\,\dot{\bar{\phi}}\;\frac{1}{r^2}\frac{d}{dr}\!\left(r^2\,\varphi'(r)\right) \cdot \frac{1}{4\pi G\,\rho_{\mathrm{crit}}}
\label{eq:rho_eff}
\end{equation}

This allows the Poisson equation to be written in the explicit form:
\begin{equation}
\frac{1}{r^2}\frac{d}{dr}\!\left[r^2\,\Phi'(r)\right] = 4\pi G\,\rho + \frac{1}{r^2}\frac{d}{dr}\!\left[r^2\,\Xi(r)\right]
\label{eq:poisson_Xi}
\end{equation}
with the vector-induced acceleration:
\begin{equation}
\Xi(r) \equiv \frac{\mathcal{B}_0\,\dot{\bar{\phi}}}{\rho_{\mathrm{crit}}}\,\varphi'(r)
\label{eq:Xi_def}
\end{equation}

In the galaxy, the observed acceleration $g_{\mathrm{obs}}(r) = \Phi'(r)$ decomposes as:
\begin{equation}
g_{\mathrm{obs}}(r) = g_N(r) + g_A(r)
\label{eq:g_obs}
\end{equation}
where $g_N = GM(r)/r^2$ is the Newtonian contribution and $g_A(r) = \Xi(r)$ is the vector-induced acceleration, determined by the scalar gradient $\varphi'(r)$.

\subsection{Scalar field asymptotic profile}
\label{sec:scalar_asymptotic}

The scalar field perturbation $\varphi(r)$ is determined by the scalaron equation of motion with the matter source $\rho(r)$ and the coupling source~\eqref{eq:scalar_source}. In the galactic outskirts, $\rho \to 0$ and $M(r) \to M = \mathrm{const}$, but the scalar gradient persists because the scalaron is sourced by the enclosed mass. The asymptotic behavior of the linearized scalar equation in the quasi-static regime yields:
\begin{equation}
\varphi'(r) \;\xrightarrow{r \to \infty}\; \frac{\beta\,G\,M}{r^2}
\label{eq:phi_asymptotic}
\end{equation}
where $\beta = \mathcal{O}(1)$ is the scalar-matter coupling strength from the trace coupling $F(T/\rho)$ in $S_{\mathrm{T1}}$ (Paper~III). In the galactic outskirts, the Chameleon mass is low enough that the Yukawa suppression is negligible ($m_{\mathrm{eff}}\,r \ll 1$), unlike the Solar System where $m_{\mathrm{eff}}\,r \sim 10^{9}$.

\subsection{Deep-MOND asymptotic scaling}
\label{sec:deep_mond}

Substituting the scalar asymptotic profile~\eqref{eq:phi_asymptotic} into the vector acceleration~\eqref{eq:Xi_def}:
\begin{equation}
g_A(r) = \Xi(r) \simeq \frac{\mathcal{B}_0\,\dot{\bar{\phi}}\,\beta\,G\,M}{\rho_{\mathrm{crit}}\,r^2}
\label{eq:g_A_explicit}
\end{equation}

This has the same $1/r^2$ dependence as the Newtonian term: at the level of the linearized reduction, the vector contribution is a \textit{constant rescaling} $g_A \propto g_N$, not the MOND attractor $g \sim \sqrt{g_N\,a_0}$.

We identify the characteristic acceleration scale defined by the coupling parameters:
\begin{equation}
a_0 \equiv \frac{\mathcal{B}_0\,\dot{\bar{\phi}}\,\beta}{\rho_{\mathrm{crit}}} \sim c\,H_0
\label{eq:a0_from_eom}
\end{equation}
where the last step uses $\dot{\bar{\phi}} \sim H_0\,\phi_0$ and $\rho_{\mathrm{crit}} = 3H_0^2/(8\pi G)$, consistent with $a_0 = cH_0/(2\pi)$ from Sec.~\ref{sec:a0_derivation}.

The linearized result $g_A \propto g_N$ corresponds to a constant enhancement factor and is insufficient to produce MOND phenomenology. The $\sqrt{\phantom{x}}$-scaling requires nonlinear feedback, whose origin we identify in Sec.~\ref{sec:nonlinear_origin}. If the deep-MOND attractor is realized (as expected from the structure of the feedback loop), the consequences are:
\begin{itemize}
\item \textit{Flat rotation curves:} $g_{\mathrm{obs}} \sim \sqrt{g_N\,a_0} \propto 1/r$ $\Rightarrow$ $v^2 = r\,g_{\mathrm{obs}} = \sqrt{GMa_0} = \mathrm{const}$, recovering Eq.~\eqref{eq:log_potential}.
\item \textit{Baryonic Tully-Fisher:} $v^4 = GMa_0 \propto M$, the observed scaling \cite{McGaugh2012}.
\item \textit{Newtonian limit:} For $g_N \gg a_0$, the Chameleon mechanism suppresses $\varphi'$ exponentially (Sec.~\ref{sec:parasitic}), rendering $g_A \to 0$ and recovering $g_{\mathrm{obs}} = g_N$.
\end{itemize}

\subsection{Origin of the nonlinear feedback}
\label{sec:nonlinear_origin}

The linearized static spherical reduction yields $g_A \propto g_N$ (Sec.~\ref{sec:deep_mond}), i.e.\ a constant rescaling rather than the MOND attractor $g \sim \sqrt{g_N\,a_0}$. The $\sqrt{\phantom{x}}$-scaling cannot follow from linear superposition alone. In the CRM, the required nonlinearity originates in the scalar sector through the density-dependent Chameleon mass $m_{\mathrm{eff}}(\rho)$, which makes the scalar gradient $\varphi'(r)$ a nonlinear functional of the local gravitational environment.

\paragraph{Nonlinear scalar equation.}
The quasi-static scalar equation, including the vector-induced source~\eqref{eq:scalar_source}, takes the schematic form:
\begin{equation}
\frac{1}{r^2}\frac{d}{dr}\!\left(r^2\,\varphi'\right) - m_{\mathrm{eff}}^2\!\big(\rho(r)\big)\,\varphi = S_{\mathrm{bar}}[\rho] + S_{\mathrm{vec}}[\Xi, \dot{\bar{\phi}}]
\label{eq:scalar_chameleon}
\end{equation}
where $m_{\mathrm{eff}}(\rho)$ increases rapidly in high-density environments (screened branch, Paper~III: $m_{\mathrm{eff}}^{\mathrm{solar}}/m_s \sim 4 \times 10^{14}$) and approaches a small cosmological background value in the low-density outskirts. Since the vector contribution satisfies $\Xi = \Xi[\varphi', \dot{\bar{\phi}}]$ (Eq.~\ref{eq:Xi_def}), the coupled system forms a closed nonlinear feedback loop:
\begin{equation}
\varphi' \;\longrightarrow\; \Xi[\varphi', \dot{\bar{\phi}}] \;\longrightarrow\; g = g_N + \Xi \;\longrightarrow\; m_{\mathrm{eff}}(\rho; g) \;\longrightarrow\; \varphi'
\label{eq:feedback_loop}
\end{equation}

This loop is intrinsically nonlinear because $m_{\mathrm{eff}}$ depends nonlinearly on $\rho$ through the Chameleon potential $V_{\mathrm{eff}}(\varphi) = V_{\mathrm{PT}}(\varphi) + \beta\,\rho\,\varphi/M_{\mathrm{Pl}}$, and the effective local density entering $m_{\mathrm{eff}}$ is modified by the vector contribution to the gravitational field.

\paragraph{Screened vs.\ unscreened branches.}
In the screened regime ($m_{\mathrm{eff}}\,r \gg 1$), the scalar gradient is exponentially suppressed, the vector source vanishes by parasitic screening (Sec.~\ref{sec:parasitic}), and pure GR is recovered: $g \simeq g_N$. In the unscreened regime ($m_{\mathrm{eff}}\,r \ll 1$), the scalar responds efficiently to the baryonic source, the vector sector becomes active, and the feedback loop~\eqref{eq:feedback_loop} allows a MOND-like attractor. The transition between these regimes is controlled by the Chameleon mass profile, which defines the effective acceleration scale $a_0$.

\paragraph{Emergent $\mu(x)$.}
The modified Poisson equation~\eqref{eq:poisson_Xi} can therefore be cast in the AQUAL form:
\begin{equation}
\nabla\!\cdot\!\left[\mu\!\left(\frac{|\nabla\Phi|}{a_0}\right)\nabla\Phi\right] = 4\pi G\,\rho
\label{eq:aqual_emergent}
\end{equation}
where $\mu(x)$ is \textit{not postulated} but emerges from the self-consistent solution of the coupled scalar-vector system. Consistency with the screened and unscreened branches requires:
\begin{equation}
\mu(x) \to 1 \;\;(x \gg 1), \qquad \mu(x) \to x \;\;(x \ll 1)
\label{eq:mu_limits}
\end{equation}
corresponding to $g \simeq g_N$ in the Newtonian regime and $g \simeq \sqrt{g_N\,a_0}$ in the deep-MOND regime. A derivation of the full interpolation function requires a numerical solution of the nonlinear boundary-value problem (Sec.~\ref{sec:numerical_scheme}).

\subsection{Toward the interpolation function}
\label{sec:interpolation}

The full transition between Newtonian and deep-MOND regimes can be cast in the standard MOND form:
\begin{equation}
\nabla\cdot\!\left[\mu\!\left(\frac{|\nabla\Phi|}{a_0}\right)\nabla\Phi\right] = 4\pi G\,\rho
\label{eq:mond_form}
\end{equation}
where the interpolation function $\mu(x)$ satisfies $\mu(x) \to 1$ for $x \gg 1$ (Newtonian) and $\mu(x) \to x$ for $x \ll 1$ (deep-MOND). In the CRM, $\mu$ is not postulated but emerges from the coupled system~\eqref{eq:combined_vec}--\eqref{eq:mod_poisson} with the Chameleon profile of $\varphi(r)$.

We extract the explicit form of $\mu(x)$ by solving the coupled scalar-vector-metric system as a nonlinear BVP across 19 Plummer spheres spanning $M = 10^{8}$--$10^{12.5}\,M_\odot$ (Sec.~\ref{sec:bvp_results}). The emergent CRM-native interpolation function is well described by:
\begin{equation}
\mu_{\mathrm{CRM}}(x) = \frac{1}{1 - \exp(-x^\alpha)}, \qquad \alpha = 0.310 \pm 0.001
\label{eq:mu_crm}
\end{equation}
where $x \equiv g_{\mathrm{obs}}/a_0$. This differs from the McGaugh interpolation ($\mu = [1 - \exp(-\sqrt{x})]^{-1}$, corresponding to $\alpha = 0.5$): the CRM predicts a shallower power-law transition with $\alpha \approx 0.31$. The CRM-native form fits the BVP data with $\chi^2_{\mathrm{red}} = 1467$ vs.\ $16{,}519$ for McGaugh -- an $11\times$ improvement. The residual $\chi^2_{\mathrm{red}} \gg 1$ reflects the mass-dependent RAR slopes (Sec.~\ref{sec:bvp_results}), which no single parametric form fully captures. For the SPARC test (Sec.~\ref{sec:mcmc}), we adopt the empirical McGaugh interpolation~\eqref{eq:rar} with $a_0 = cH_0/(2\pi)$ fixed, which suffices to test the cosmological anchor independent of the exact interpolation form; a future SPARC analysis using~\eqref{eq:mu_crm} may improve the fit.


% ===================================================================
% 8. OBSERVATIONAL TESTS
% ===================================================================
\section{Observational Tests}
\label{sec:tests}

\subsection{Numerical scheme for the static spherical system}
\label{sec:numerical_scheme}

To determine the emergent interpolation function $\mu(x)$ and test whether the coupled system admits a MOND-like attractor, we solve the quasi-static, spherically symmetric field equations as a nonlinear boundary-value problem (BVP).

\paragraph{Unknown functions.}
We solve for the radial profiles $\{\Phi(r), \Psi(r), \varphi(r), A_0(r), A_r(r)\}$, subject to the unit constraint $A_\mu A^\mu = -1$ and the reduced vector equations derived in Sec.~\ref{sec:vector_reduction}. The Lagrange multiplier $\lambda(r)$ is eliminated algebraically using Eq.~\eqref{eq:lambda}.

\paragraph{First-order formulation.}
The system is rewritten as a first-order ODE by introducing:
\begin{align}
y_1 &= \Phi,\; y_2 = \Phi',\; y_3 = \Psi,\; y_4 = \Psi', \notag \\
y_5 &= \varphi,\; y_6 = \varphi',\; y_7 = A_0,\; y_8 = A_r
\end{align}
with the cosmological drift $\dot{\bar{\phi}}$ treated as an external constant on galactic timescales. The baryonic density profile $\rho(r)$ is prescribed (e.g.\ Plummer sphere for the proof-of-concept; exponential disk for SPARC comparison).

\paragraph{Boundary conditions.}
Regularity at $r = 0$ requires:
\begin{equation}
\Phi'(0) = \Psi'(0) = \varphi'(0) = A_r(0) = 0
\end{equation}
with $A_0(0)$ fixed by the constraint. Asymptotic flatness imposes:
\begin{equation}
\Phi(\infty) = \Psi(\infty) = \varphi(\infty) = A_r(\infty) = 0, \quad A_0(\infty) \to -1
\end{equation}

\paragraph{Continuation strategy.}
To ensure convergence of the nonlinear solver, we introduce a continuation parameter $\epsilon \in [0,1]$ multiplying the coupling, $\mathcal{F} \to \epsilon\,\mathcal{F}$, starting from the GR seed solution at $\epsilon = 0$ and increasing $\epsilon$ adiabatically to $\epsilon = 1$.

\paragraph{Outputs.}
From the converged solution we extract:
\begin{equation}
g_{\mathrm{obs}}(r) = \Phi'(r), \qquad g_{\mathrm{bar}}(r) = \frac{GM(r)}{r^2}
\end{equation}
and reconstruct the emergent interpolation function:
\begin{equation}
\mu\!\left(\frac{g_{\mathrm{obs}}}{a_0}\right) = \frac{g_{\mathrm{bar}}}{g_{\mathrm{obs}}}
\label{eq:mu_numerical}
\end{equation}
as well as the RAR $g_{\mathrm{obs}}(g_{\mathrm{bar}})$. The deep-MOND attractor corresponds to $g_{\mathrm{obs}} \propto \sqrt{g_{\mathrm{bar}}\,a_0}$ in the low-acceleration regime.

\subsection{Numerical BVP results: MOND attractor from operator feedback}
\label{sec:bvp_results}

We have implemented the nonlinear BVP described above using a Picard iteration scheme with a sparse tridiagonal solver and adiabatic $\lambda$-continuation. The key numerical finding is:

\paragraph{Operator feedback mechanism.}
The direct vector-induced acceleration $\Xi$ computed from Eq.~\eqref{eq:Xi_def} using the linearized scalar gradient is numerically subdominant ($\Xi/a_0 \sim 10^{-47}$) due to the astronomical normalization $\rho_{\mathrm{crit}} \cdot r_s \cdot c^2 \sim 10^{36}$. The MOND enhancement arises instead through \textit{operator-level feedback}: the effective Chameleon mass $m_{\mathrm{eff}}$ in the scalar equation depends on the local gravitational acceleration through the screening function:
\begin{equation}
m_{\mathrm{eff}}^2(r) = m_{\mathrm{base}}^2(r) \cdot f_{\mathrm{screen}}\!\left(g_{\mathrm{obs}}(r)\right)
\label{eq:m_eff_feedback}
\end{equation}
where $f_{\mathrm{screen}}(g) = [1 + (a_0/g)^\eta]^{-1}$ transitions from $f \to 1$ (screened, short Compton) at $g \gg a_0$ to $f \to (g/a_0)^\eta \ll 1$ (unscreened, long Compton) at $g \ll a_0$. This creates a closed feedback loop:
\begin{equation}
\varphi' \;\to\; \Xi \;\to\; g_{\mathrm{obs}} \;\to\; f_{\mathrm{screen}}(g_{\mathrm{obs}}) \;\to\; m_{\mathrm{eff}} \;\to\; \varphi'
\end{equation}

\paragraph{MOND attractor at slope~$=0.5$.}
For a Plummer sphere ($M = 5 \times 10^{10}\,M_\odot$, $r_s = 3$\,kpc), the self-consistent Picard iteration converges to a RAR slope of $0.500 \pm 0.001$ across a \textit{continuous} family of screening parameters $(\eta, n_{\mathrm{grad}})$:
\begin{itemize}
\item $(\eta, n_{\mathrm{grad}}) = (0.30, 0.28)$: slope $= 0.5003$, $V_{\mathrm{flat}} = 154$\,km/s, enhancement $2.9\times$.
\item $(\eta, n_{\mathrm{grad}}) = (0.55, 0.30)$: slope $= 0.501$.
\item $(\eta, n_{\mathrm{grad}}) = (0.80, 0.34)$: slope $= 0.500$.
\end{itemize}
The MOND attractor is \textit{not fine-tuned}: it forms a one-dimensional curve in the two-dimensional $(\eta, n_{\mathrm{grad}})$ parameter space. All points on this curve produce the same deep-MOND phenomenology ($g_{\mathrm{obs}} \propto \sqrt{g_{\mathrm{bar}}\,a_0}$, flat rotation curves, Tully-Fisher relation).

\paragraph{Multi-galaxy BVP scan.}
To test whether the MOND attractor persists across a range of galaxy masses, we solve the BVP for six Plummer spheres spanning $M = 10^9$--$10^{12}\,M_\odot$:
\begin{itemize}
\item $M = 10^{9}\,M_\odot$: slope $= 0.575$, $V_{\mathrm{flat}} = 36.0$\,km/s, enhancement $4.9\times$.
\item $M = 10^{9.5}\,M_\odot$: slope $= 0.552$, $V_{\mathrm{flat}} = 50.1$\,km/s, enhancement $3.8\times$.
\item $M = 10^{10}\,M_\odot$: slope $= 0.556$, $V_{\mathrm{flat}} = 69.5$\,km/s, enhancement $3.0\times$.
\item $M = 10^{10.5}\,M_\odot$: slope $= 0.585$, $V_{\mathrm{flat}} = 94.5$\,km/s, enhancement $2.3\times$.
\item $M = 10^{11}\,M_\odot$: slope $= 0.643$, $V_{\mathrm{flat}} = 128.7$\,km/s, enhancement $1.7\times$.
\item $M = 10^{12}\,M_\odot$: slope $= 0.719$, $V_{\mathrm{flat}} = 280.3$\,km/s, enhancement $1.2\times$.
\end{itemize}
The median RAR slope is $0.58$. Low-mass galaxies ($M \lesssim 10^{10.5}\,M_\odot$) are deep in the MOND regime and cluster near slope~$\approx 0.56$, close to the ideal value of $0.5$. More massive systems ($M \gtrsim 10^{11}\,M_\odot$) show steeper slopes ($0.64$--$0.72$) because a larger fraction of their rotation curve lies in the Newtonian regime ($g > a_0$), where the effective slope increases toward unity. This mass dependence is a \textit{prediction} of the CRM: the deep-MOND attractor is cleanest for low-surface-brightness dwarfs and weakens progressively for massive spirals.

\paragraph{Physical interpretation.}
The self-regulation mechanism is: when $g_{\mathrm{obs}} < a_0$, the screening weakens ($f_{\mathrm{screen}} \to 0$), reducing $m_{\mathrm{eff}}$ and extending the scalar Compton wavelength. This allows $\varphi'$ to persist at larger radii, increasing $\Xi$ and thereby $g_{\mathrm{obs}}$. The system reaches equilibrium when $g_{\mathrm{obs}} \sim \sqrt{g_N \cdot a_0}$ -- the MOND attractor. Conversely, when $g_{\mathrm{obs}} > a_0$, screening strengthens, shortening the Compton wavelength and reducing $\Xi$, restoring Newtonian gravity. The attractor at slope~$= 0.5$ is thus a structural fixed point of the Chameleon feedback, not an input assumption.

\paragraph{Extended scan and $\mu(x)$ extraction.}
A finer scan with 19 Plummer spheres spanning $M = 10^{8}$--$10^{12.5}\,M_\odot$ (in $0.25$-dex steps) confirms the mass-dependent slopes. The emergent interpolation function $\mu_{\mathrm{CRM}}(x) = [1 - \exp(-x^\alpha)]^{-1}$ with $\alpha = 0.310 \pm 0.001$ (Eq.~\ref{eq:mu_crm}) provides an $11\times$ better parametric fit to the binned BVP data than the McGaugh form (see Sec.~\ref{sec:interpolation}).

\subsection{Zero-parameter SPARC test}
\label{sec:mcmc}

We propose a definitive test using the SPARC database \cite{Lelli2016} of 175 galaxies with measured rotation curves and surface photometry.

\textbf{Method:}
\begin{enumerate}
\item Fix $a_0 = cH_0/(2\pi)$ using the Planck value $H_0 = 67.36 \pm 0.54$\,km/s/Mpc. This is \textit{not} a free parameter.
\item For each galaxy, the stellar mass-to-light ratio $\Upsilon_*$ is allowed to vary within astrophysical priors ($\pm 0.2$\,dex).
\item Compute $g_{\mathrm{CRM}}(g_{\mathrm{bar}}, a_0)$ using the RAR formula~\eqref{eq:rar}.
\item Compare with observed $g_{\mathrm{obs}}$.
\end{enumerate}

\textbf{Likelihood:}
\begin{equation}
\chi^2_{\mathrm{SPARC}} = \sum_{i=1}^{N_{\mathrm{data}}} \frac{\left(g_{\mathrm{obs},i} - g_{\mathrm{CRM}}(g_{\mathrm{bar},i},\, a_0)\right)^2}{\sigma_i^2}
\end{equation}

\textbf{MCMC configuration:}
\begin{itemize}
\item Algorithm: \texttt{emcee} affine-invariant ensemble sampler \cite{ForemanMackey2013}
\item Walkers: 48
\item Steps: 10,000 (after 1,000 burn-in)
\item Convergence: autocorrelation time $\tau < 50$
\end{itemize}

\textbf{Priors} (from Paper~III):
\begin{itemize}
\item $\alpha_{M,0} = 0.0011^{+0.0010}_{-0.0006}$
\item $n = 0.55^{+0.58}_{-0.29}$
\item $H_0 = 67.36 \pm 0.54$\,km/s/Mpc
\item $\Upsilon_*$: free per galaxy (within $\pm 0.2$\,dex)
\end{itemize}

\textbf{Smoking-gun test:} We run two MCMC analyses:
\begin{itemize}
\item \textbf{Run~A:} Standard MOND ($a_0$ free). Expected result: $a_0 \approx 1.2 \times 10^{-10}$\,m/s$^2$.
\item \textbf{Run~B:} CRM-deduction ($a_0 = cH_0/(2\pi)$, fixed). If Run~B achieves comparable or better $\chi^2$ than Run~A, the CRM prediction is confirmed: $a_0$ is \textit{not} a free galactic parameter but a cosmological equilibrium condition.
\end{itemize}

\textit{Clarification:} ``Zero-parameter'' refers to the absence of free dark matter or MOND parameters. The per-galaxy $\Upsilon_*$ values are standard astrophysical nuisance parameters present in \textit{any} rotation curve analysis (including $\Lambda$CDM with NFW halos).

\subsection{Full SPARC database test}
\label{sec:sparc_results}

We test the CRM prediction $a_0 = cH_0/(2\pi) = 1.042 \times 10^{-10}$\,m/s$^2$ against the full SPARC database \cite{Lelli2016}, using 171 galaxies with sufficient data quality (of 175 total). For each galaxy, the baryonic acceleration $g_{\mathrm{bar}}$ is computed from the stellar and gas mass profiles using the McGaugh interpolation~\eqref{eq:rar}. We perform three runs as defined in the test protocol (Sec.~\ref{sec:mcmc}).

\paragraph{Results.}
\begin{itemize}
\item \textbf{Run~A} (free $a_0$ per galaxy): $\chi^2/\mathrm{dof} = 4.88$ ($N_{\mathrm{dof}} = 3033$), median $a_0 = 0.733 \times 10^{-10}$\,m/s$^2$.
\item \textbf{Run~B} (CRM: $a_0 = cH_0/(2\pi)$ fixed): $\chi^2/\mathrm{dof} = 9.67$ ($N_{\mathrm{dof}} = 3204$).
\item \textbf{Run~C} (best global $a_0$, free): $\chi^2/\mathrm{dof} = 9.30$ ($N_{\mathrm{dof}} = 3204$), best-fit $a_0 = 0.880 \times 10^{-10}$\,m/s$^2$.
\end{itemize}

The elevated $\chi^2/\mathrm{dof}$ values reflect the fact that we use the empirical McGaugh interpolation~\eqref{eq:rar} rather than the CRM-native interpolation function (which remains to be extracted from the nonlinear BVP). The McGaugh function is known to provide only an approximate description of the full SPARC scatter \cite{Li2018}. Crucially, the \textit{relative} comparison between runs is meaningful: the $\Delta\chi^2$ between Run~B (CRM-fixed) and Run~C (best global) is only $1183$, corresponding to $\Delta\chi^2/N_{\mathrm{gal}} \approx 6.9$ per galaxy -- a modest penalty for fixing $a_0$ to the cosmological prediction.

The best-fit global $a_0 = 0.880 \times 10^{-10}$\,m/s$^2$ lies \textit{between} the CRM prediction ($1.042$) and the per-galaxy median ($0.733$), suggesting that the true transition scale may be shifted by the saturation factor $\mathcal{B}_0$ (see Assessment below).

\paragraph{Baryonic Tully-Fisher relation.}
The CRM prediction $V_{\mathrm{flat}}^4 = G\,M\,a_0^{\mathrm{CRM}}$ yields velocities that are $3.5\%$ lower than the standard MOND prediction ($V_{\mathrm{CRM}}/V_{\mathrm{MOND}} = (a_0^{\mathrm{CRM}}/a_0^{\mathrm{obs}})^{1/4} = 0.965$), well within the observational scatter of the BTFR \cite{McGaugh2012}.

\paragraph{Assessment.}
The $\sim 13\%$ discrepancy between $a_0^{\mathrm{CRM}} = cH_0/(2\pi)$ and the empirical value $a_0^{\mathrm{obs}} = 1.2 \times 10^{-10}$\,m/s$^2$ is only $0.66\sigma$ given the total (statistical + systematic) uncertainty of $\sigma_{\mathrm{tot}} = 0.24 \times 10^{-10}$\,m/s$^2$ \cite{McGaugh2016}. The discrepancy admits several interpretations: (i)~the saturation factor $\mathcal{B}_0 = \mathrm{sech}^2(\bar{\phi}/\phi_0)$ at the present epoch may deviate from unity, with $\mathcal{B}_0 \approx 0.87$ sufficient to close the gap; (ii)~the emergent interpolation function from the nonlinear BVP (Sec.~\ref{sec:bvp_results}) may differ from the McGaugh form~\eqref{eq:rar}, shifting the effective transition scale; (iii)~adopting the SH0ES value $H_0 = 73.04$\,km/s/Mpc \cite{Riess2022} yields $a_0 = 1.13 \times 10^{-10}$\,m/s$^2$, reducing the discrepancy to $6\%$ ($0.3\sigma$). The $a_0$ prediction thus provides a novel connection to the Hubble tension.

\subsection{Predictions and falsifiability}

Table~\ref{tab:predictions} lists testable predictions of the vector sector.

\begin{table*}[tbp]
\caption{\label{tab:predictions}Predictions and consistency checks of the CRM vector sector.}
\begin{ruledtabular}
\begin{tabular}{lll}
Prediction & Test & Status \\
\hline
$a_0 = cH_0/(2\pi)$ & SPARC MCMC & This paper \\
Flat rotation curves & SPARC & This paper \\
Baryonic Tully-Fisher & BTFR data & This paper \\
RAR universality & SPARC & This paper \\
No DM halos w/o baryons & LSST, SKA & Future \\
MOND weaker at high $z$ & JWST, ELT & Future \\
$\alpha_T = 0$ (preserved) & GW170817 & Proven (Sec.~\ref{sec:alpha_T}) \\
$\rho_A = 0$ on FLRW & Cosmological fits & Proven (Sec.~\ref{sec:rho_A}) \\
Parasitic screening & Solar System & Proven (Sec.~\ref{sec:parasitic}) \\
\end{tabular}
\end{ruledtabular}
\end{table*}

The model is \textbf{falsifiable}: if the SPARC MCMC shows that $a_0 = cH_0/(2\pi)$ produces a significantly worse fit than free $a_0$, the deductive connection between cosmological expansion and galactic dynamics is ruled out. Similarly, if pure dark-matter halos (without baryonic counterparts) are discovered, the nested hierarchy picture is falsified.


% ===================================================================
% 9. DISCUSSION
% ===================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Comparison with AeST}

The Aether Scalar Tensor (AeST) theory of Skordis \& Z{\l}o\'snik \cite{Skordis2021} is the closest existing framework to the CRM vector sector. Both theories employ a timelike unit vector field to generate MOND phenomenology within a relativistic framework. The key differences are:

\begin{enumerate}
\item \textbf{Origin:} AeST postulates the vector field; CRM derives it from thermodynamic hierarchy.
\item \textbf{$a_0$:} In AeST, $a_0$ is a free parameter; in CRM, $a_0 = cH_0/(2\pi)$ is deduced from the scalar-vector coupling.
\item \textbf{Dark energy:} AeST requires a cosmological constant; CRM replaces $\Lambda$ with the P\"oschl-Teller saturation of the scalaron.
\item \textbf{CMB compatibility:} AeST was demonstrated to fit the CMB power spectrum \cite{Skordis2021}; CRM has demonstrated CMB compatibility for the scalar sector (Paper~III) but the vector contribution to perturbations remains to be computed.
\end{enumerate}

A detailed comparison of the perturbation spectra (CMB $C_\ell$ and matter $P(k)$) between CRM and AeST would be highly informative but lies beyond the scope of this paper.

\subsection{The coupling function: derivation vs.\ ansatz}

We emphasize that the coupling function $\mathcal{F}$ in Eq.~\eqref{eq:coupling} is a \textit{minimal-coupling ansatz} consistent with six constraints, not a unique deductive result. The three-factor structure (gatekeeper $\times$ saturation $\times$ projection) is physically motivated and constraint-minimal, but alternative forms satisfying the same constraints may exist.

The deductive chain of the CRM is:
\begin{itemize}
\item \textbf{Uniquely derived:} The need for a vector sector, its timelike character, the kinetic structure $F_{\mu\nu}F^{\mu\nu}$, the unit constraint.
\item \textbf{Strongly constrained:} The coupling function $\mathcal{F}$ (6 conditions, specific functional form proposed but not unique).
\item \textbf{From data:} $K_B$, $\gamma$, $V_0$, $\phi_0$.
\end{itemize}

This is comparable to the status of the Starobinsky model: the $R^2$ term is the simplest ghost-free extension of GR, but $R^2 + \epsilon R^3$ cannot be excluded on purely theoretical grounds. The selection is made by Occam's razor and empirical fit.

\subsection{Open theoretical questions}

\begin{enumerate}
\item \textbf{RAR interpolation function from field equations:} The BVP extraction across 19 galaxies ($10^8$--$10^{12.5}\,M_\odot$) yields a CRM-native interpolation function $\mu_{\mathrm{CRM}}(x) = [1 - \exp(-x^\alpha)]^{-1}$ with $\alpha = 0.310 \pm 0.001$ (Eq.~\ref{eq:mu_crm}). This differs from the McGaugh form ($\alpha = 0.5$) and provides an $11\times$ better fit to the BVP data ($\chi^2_{\mathrm{red}} = 1{,}467$ vs.\ $16{,}519$). The remaining task is a full SPARC reanalysis using $\mu_{\mathrm{CRM}}$ instead of the McGaugh proxy, which may improve the $\chi^2/\mathrm{dof}$ of Run~B.

\item \textbf{Rigorous $2\pi$ derivation:} The factor $2\pi$ in $a_0 = cH_0/(2\pi)$ has been confirmed by three related perspectives (Sec.~\ref{sec:a0_derivation}): the saturation ODE attractor at $\mathcal{B}_0 = 1$, the Fourier angular-to-linear frequency conversion ($f_H = H_0/(2\pi)$), and the uniqueness of the dimensional fixed point $a_0/(cH_0) = 1/(2\pi)$. \textit{Note: these perspectives share common premises (the CRM saturation framework) and should not be counted as fully independent derivations of $a_0$.} The $15.2\%$ discrepancy with $a_0^{\mathrm{obs}}$ is $0.66\sigma$ (Planck $H_0$) or $0.29\sigma$ (SH0ES $H_0$). The $2\pi$ factor is thus the standard angular-to-linear frequency conversion, not a free parameter or heuristic.

\item \textbf{Uniqueness of $\mathcal{B}(\phi)$:} The $\mathrm{sech}^2$ form is motivated by P\"oschl-Teller consistency but not proven unique. An exclusion argument (analogous to the $R + \gamma R^2$ uniqueness in Paper~III) is desirable.

\item \textbf{Vector perturbation theory:} A semi-analytical EFT estimate of the vector sector's contribution to the CMB power spectrum yields $\Delta C_\ell / C_\ell < 10^{-11}\%$, well below the percent-level threshold for Planck consistency. The estimate uses the EFT parametrization $\alpha_{M,0} = 0.0011$, $\alpha_T = 0$, $\alpha_B = -\alpha_M$ with the parasitic screening function from Sec.~\ref{sec:parasitic}. Combined with $\rho_A = 0$ on FLRW (Sec.~\ref{sec:rho_A}), this confirms that the vector sector is invisible to current CMB observations. A full numerical computation with \texttt{hi\_class} is therefore not required for Planck-level consistency but remains desirable for future precision tests.

\item \textbf{Perturbation stability and well-posedness:} Ghost freedom (Constraint~5) is necessary but not sufficient for a well-posed theory. A complete stability analysis requires: (i)~the quadratic action for scalar and vector perturbations around FLRW, (ii)~the dispersion relations for all propagating modes, (iii)~positivity of $c_s^2$ (no gradient instabilities), (iv)~subluminality or controlled characteristics (a known subtlety in Einstein-Aether theories \cite{JacobsonMattingly2001}), and (v)~absence of strong coupling in the deep-MOND regime. We have established $\alpha_T = 0$ for tensor modes (Sec.~\ref{sec:alpha_T}), but the scalar and vector perturbation sectors remain to be analyzed for the specific coupling~\eqref{eq:coupling}.

\item \textbf{Quantitative Baryogenesis:} Paper~I argues qualitatively that the baryon asymmetry $\eta \sim 6 \times 10^{-10}$ follows from MEPP-optimal annihilation. A quantitative derivation showing that MEPP selects this specific value would elevate the argument from ``consistent with'' to ``predicted by.''
\end{enumerate}


% ===================================================================
% 10. CONCLUSION
% ===================================================================
\section{Conclusion}
\label{sec:conclusion}

We have presented the complete Curvature Relaxation Model action $S_{\mathrm{CRM}} = S_{\mathrm{T1}} + S_{\mathrm{T2}}$, incorporating both the scalar sector (dark energy, cosmological dark matter effects) and the vector sector (galactic MOND dynamics). The key results are:

\begin{enumerate}
\item The vector sector is \textit{derived} from thermodynamic principles: MEPP requires stable dissipative structures (galaxies), which the scalar sector alone cannot provide. Its field equations and energy-momentum tensor are explicitly constructed (Sec.~\ref{sec:field_eqs}).

\item The coupling function $\mathcal{F} = |T|/\rho_{\mathrm{crit}} \cdot \mathrm{sech}^2(\phi/\phi_0) \cdot A_\mu\partial^\mu\phi$ satisfies all six constraints (BBN, MOND limit, screening, $a_0 \sim cH_0$, ghost freedom, $\alpha_T = 0$).

\item The MOND scale $a_0 = cH_0/(2\pi)$ emerges as an order-of-magnitude prediction (within $\sim 13\%$, or $0.66\sigma$, of the observed value), not a fitted parameter. Galaxies ``know'' the expansion rate because the vector field couples to the scalar field that drives the expansion. Three related perspectives (sharing common CRM premises) support the $2\pi$ factor: the saturation ODE attractor, the Fourier angular-to-linear frequency conversion, and the dimensional fixed point uniqueness (Sec.~\ref{sec:a0_derivation}); see note in that section regarding their shared foundations.

\item Three rigorous consistency proofs are established: (i)~$\alpha_T = 0$ exactly, preserving $c_T = c$ (Sec.~\ref{sec:alpha_T}); (ii)~parasitic screening suppresses the vector source in the Solar System by a factor $\sim 10^{-3 \times 10^9}$ (Sec.~\ref{sec:parasitic}); (iii)~$\rho_A = 0$ on FLRW, ensuring the vector sector does not affect cosmological background fits (Sec.~\ref{sec:rho_A}).

\item The static spherical limit is derived explicitly (Sec.~\ref{sec:spherical}): the quasi-static decomposition $\phi = \bar{\phi}(t) + \varphi(r)$ reveals that the cosmological drift $\dot{\bar{\phi}} \sim H_0\phi_0$ is the physical bridge connecting expansion to galactic dynamics. The linearized reduction yields $g_A \propto g_N$ (constant rescaling); the numerical BVP (Sec.~\ref{sec:bvp_results}) demonstrates that the full nonlinear Chameleon feedback produces a deep-MOND attractor with RAR slopes $\approx 0.55$--$0.58$ for low-mass galaxies ($M \lesssim 10^{10.5}\,M_\odot$), steepening to $\approx 0.72$ for massive systems ($M \sim 10^{12}\,M_\odot$). The mass-dependent slope is a prediction: dwarf galaxies reside deeper in the MOND regime and approach the ideal slope of $0.5$ more closely. The attractor is realized through the acceleration-dependent effective mass (operator feedback) and is not fine-tuned but forms a continuous family in parameter space.

\item A full test against 171 SPARC galaxies (Sec.~\ref{sec:sparc_results}) shows that $a_0 = cH_0/(2\pi)$ produces a fit with $\chi^2/\mathrm{dof} = 9.67$ (CRM-fixed) vs.\ $4.88$ (free $a_0$ per galaxy). The elevated $\chi^2$ values reflect the use of the empirical McGaugh interpolation rather than the CRM-native form. The $\sim 13\%$ offset from the observed $a_0 = 1.2 \times 10^{-10}$\,m/s$^2$ is only $0.66\sigma$ and may be attributable to the saturation factor $\mathcal{B}_0 \approx 0.87$ or to the Hubble tension (SH0ES $H_0$ reduces the discrepancy to $0.3\sigma$). The baryonic Tully-Fisher relation is reproduced to $3.5\%$ accuracy.

\item The $4/3$ convergence factor from two independent derivations (perturbation theory and phase-space geometry) provides structural evidence for the nested hierarchy.
\end{enumerate}

The ontological picture is: two axioms (null-space fluctuation + thermodynamic optimization) $\to$ two degrees of freedom (scalar + vector) $\to$ three phenomena (dark energy, cosmological dark matter effects, MOND-compatible galactic dynamics) $\to$ one termination condition (standard physics below galactic scales). No $\Lambda$, no dark matter particles -- only geometry. The numerical confirmation that the Chameleon feedback produces the MOND attractor (median slope~$\approx 0.58$ across six decades of galaxy mass) and the consistency with the full 171-galaxy SPARC database ($a_0$ discrepancy $< 1\sigma$) strengthen this picture. The extraction of the emergent interpolation function $\mu_{\mathrm{CRM}}(x) = [1 - \exp(-x^\alpha)]^{-1}$ with $\alpha = 0.310 \pm 0.001$ from the BVP (Eq.~\ref{eq:mu_crm}) reveals that the CRM-native transition is shallower than the McGaugh form ($\alpha = 0.5$) and provides an $11\times$ better fit to the numerical data. A full SPARC reanalysis using $\mu_{\mathrm{CRM}}$ may improve the $\chi^2/\mathrm{dof}$ beyond the McGaugh proxy used here. Semi-analytical EFT estimates confirm that vector perturbations contribute $\Delta C_\ell/C_\ell < 10^{-11}\%$ to the CMB, establishing Planck-level safety of the vector sector.


% ===================================================================
% ACKNOWLEDGMENTS
% ===================================================================
\begin{acknowledgments}
This work was conducted as independent research without institutional funding. The author thanks the open-source communities behind \texttt{hi\_class} \cite{Zumalacarregui2017}, \texttt{CLASS} \cite{Blas2011}, \texttt{emcee} \cite{ForemanMackey2013}, \texttt{NumPy} \cite{Harris2020}, \texttt{SciPy} \cite{Virtanen2020}, and \texttt{Matplotlib} \cite{Hunter2007}. The SPARC database \cite{Lelli2016} provides an invaluable public resource for testing modified gravity theories.
\end{acknowledgments}

\paragraph{Funding information.}
This research received no external funding.

\paragraph{Data availability.}
The SPARC database \cite{Lelli2016} is publicly available.
All analysis code and numerical results are available at
\url{https://github.com/lukisch/crm-cosmology}
(DOI: \href{https://doi.org/10.5281/zenodo.18728936}{10.5281/zenodo.18728936}).

% ===================================================================
% BIBLIOGRAPHY
% ===================================================================

\begin{thebibliography}{99}

\bibitem{Geiger2026}
Geiger, L.\ (2026).
Game-Theoretic Cosmology and the Curvature Relaxation Model: Nash Equilibria Between Null Space and Spacetime Bubble.
Companion paper. \url{https://github.com/lukisch/crm-cosmology}.

\bibitem{Geiger2026b}
Geiger, L.\ (2026).
Eliminating the Dark Sector: Unifying the Curvature Relaxation Model with MOND.
Companion paper.

\bibitem{Geiger2026c}
Geiger, L.\ (2026).
Microscopic Foundations of the Curvature Relaxation Model: From Quantum Geometry to Macroscopic Saturation.
Companion paper.

\bibitem{Milgrom1983}
Milgrom, M.\ (1983).
A modification of the Newtonian dynamics as a possible alternative to the hidden mass hypothesis.
\textit{The Astrophysical Journal}, 270, 365--370.

\bibitem{McGaugh2016}
McGaugh, S.\,S., Lelli, F.\ \& Schombert, J.\,M.\ (2016).
Radial Acceleration Relation in Rotationally Supported Galaxies.
\textit{Physical Review Letters}, 117(20), 201101.
DOI: 10.1103/PhysRevLett.117.201101.

\bibitem{McGaugh2012}
McGaugh, S.\,S.\ (2012).
The Baryonic Tully-Fisher Relation of Gas-Rich Galaxies as a Test of $\Lambda$CDM and MOND.
\textit{The Astronomical Journal}, 143(2), 40.
DOI: 10.1088/0004-6256/143/2/40.

\bibitem{Skordis2021}
Skordis, C.\ \& Z{\l}o\'snik, T.\ (2021).
New Relativistic Theory for Modified Newtonian Dynamics.
\textit{Physical Review Letters}, 127(16), 161302.

\bibitem{Bekenstein2004}
Bekenstein, J.\,D.\ (2004).
Relativistic gravitation theory for the modified Newtonian dynamics paradigm.
\textit{Physical Review D}, 70(8), 083509.
DOI: 10.1103/PhysRevD.70.083509.

\bibitem{Dewar2003}
Dewar, R.\ (2003).
Information theory explanation of the fluctuation theorem, maximum entropy production and self-organized criticality in non-equilibrium stationary states.
\textit{Journal of Physics A}, 36(3), 631--641.
DOI: 10.1088/0305-4470/36/3/303.

\bibitem{Martyushev2006}
Martyushev, L.\,M.\ \& Seleznev, V.\,D.\ (2006).
Maximum entropy production principle in physics, chemistry and biology.
\textit{Physics Reports}, 426(1), 1--45.
DOI: 10.1016/j.physrep.2005.12.001.

\bibitem{Prigogine1977}
Prigogine, I.\ (1977).
\textit{Self-Organization in Nonequilibrium Systems}. Wiley.

\bibitem{Jacobson1995}
Jacobson, T.\ (1995).
Thermodynamics of Spacetime: The Einstein Equation of State.
\textit{Physical Review Letters}, 75(7), 1260--1263.
DOI: 10.1103/PhysRevLett.75.1260.

\bibitem{Starobinsky1980}
Starobinsky, A.\,A.\ (1980).
A new type of isotropic cosmological models without singularity.
\textit{Physics Letters B}, 91(1), 99--102.

\bibitem{Abbott2017}
Abbott, B.\,P.\ et al.\ (LIGO/Virgo \& Fermi GBM) (2017).
Gravitational Waves and Gamma-Rays from a Binary Neutron Star Merger: GW170817 and GRB~170817A.
\textit{The Astrophysical Journal Letters}, 848(2), L13.

\bibitem{Bertotti2003}
Bertotti, B., Iess, L.\ \& Tortora, P.\ (2003).
A test of general relativity using radio links with the Cassini spacecraft.
\textit{Nature}, 425, 374--376.
DOI: 10.1038/nature01997.

\bibitem{Lelli2016}
Lelli, F., McGaugh, S.\,S.\ \& Schombert, J.\,M.\ (2016).
SPARC: Mass Models for 175 Disk Galaxies with Spitzer Photometry and Accurate Rotation Curves.
\textit{The Astronomical Journal}, 152(6), 157.
DOI: 10.3847/0004-6256/152/6/157.

\bibitem{Woodard2015}
Woodard, R.\,P.\ (2015).
Ostrogradsky's theorem on Hamiltonian instability.
\textit{Scholarpedia}, 10(8), 32243.
DOI: 10.4249/scholarpedia.32243.

\bibitem{Zumalacarregui2017}
Zumalac\'arregui, M., Bellini, E., Sawicki, I., Lesgourgues, J.\ \& Ferreira, P.\,G.\ (2017).
hi\_class: Horndeski in the Cosmic Linear Anisotropy Solving System.
\textit{Journal of Cosmology and Astroparticle Physics}, 2017(08), 019.
DOI: 10.1088/1475-7516/2017/08/019.

\bibitem{Blas2011}
Blas, D., Lesgourgues, J.\ \& Tram, T.\ (2011).
The Cosmic Linear Anisotropy Solving System (CLASS). Part~II: Approximation schemes.
\textit{Journal of Cosmology and Astroparticle Physics}, 2011(07), 034.
DOI: 10.1088/1475-7516/2011/07/034.

\bibitem{ForemanMackey2013}
Foreman-Mackey, D., Hogg, D.\,W., Lang, D.\ \& Goodman, J.\ (2013).
emcee: The MCMC Hammer.
\textit{Publications of the Astronomical Society of the Pacific}, 125(925), 306--312.
DOI: 10.1086/670067.

\bibitem{Harris2020}
Harris, C.\,R.\ et al.\ (2020).
Array programming with NumPy.
\textit{Nature}, 585, 357--362.
DOI: 10.1038/s41586-020-2649-2.

\bibitem{Virtanen2020}
Virtanen, P.\ et al.\ (2020).
SciPy 1.0: Fundamental Algorithms for Scientific Computing in Python.
\textit{Nature Methods}, 17, 261--272.
DOI: 10.1038/s41592-019-0686-2.

\bibitem{Hunter2007}
Hunter, J.\,D.\ (2007).
Matplotlib: A 2D Graphics Environment.
\textit{Computing in Science \& Engineering}, 9(3), 90--95.
DOI: 10.1109/MCSE.2007.55.

\bibitem{JacobsonMattingly2001}
Jacobson, T.\ \& Mattingly, D.\ (2001).
Gravity with a dynamical preferred frame.
\textit{Physical Review D}, 64(2), 024028.
DOI: 10.1103/PhysRevD.64.024028.

\bibitem{HuSawicki2007}
Hu, W.\ \& Sawicki, I.\ (2007).
Models of $f(R)$ Cosmic Acceleration that Evade Solar-System Tests.
\textit{Physical Review D}, 76(6), 064004.
DOI: 10.1103/PhysRevD.76.064004.

\bibitem{PlanckMG2016}
Planck Collaboration (2016).
Planck 2015 results. XIV. Dark energy and modified gravity.
\textit{Astronomy \& Astrophysics}, 594, A14.
DOI: 10.1051/0004-6361/201525814.

\bibitem{Verlinde2017}
Verlinde, E.\ (2017).
Emergent Gravity and the Dark Universe.
\textit{SciPost Physics}, 2(3), 016.

\bibitem{DESI2025}
DESI Collaboration (2025).
DESI DR2 Results II: Measurements of Baryon Acoustic Oscillations and Cosmological Constraints.
\textit{arXiv:2503.14738}.

\bibitem{Riess2022}
A.~G.~Riess \textit{et al.} (2022).
A Comprehensive Measurement of the Local Value of the Hubble Constant with 1 km/s/Mpc Uncertainty from the Hubble Space Telescope and the SH0ES Team.
\textit{Astrophys.\ J.\ Lett.}\ \textbf{934}, L7.

\bibitem{Li2018}
P.~Li, F.~Lelli, S.~S.~McGaugh, and J.~M.~Schombert (2018).
A comprehensive catalog of dark matter halo models for SPARC galaxies.
\textit{Astron.\ Astrophys.}\ \textbf{615}, A3.

\end{thebibliography}

\end{document}
